M.Sc. Proposal: \\ \bigskip Computational Electrokinetics

### 2011-12-18 notes/work/ch02.tex
Rename

The boundary conditions at the ion-exchange boundary ($r=1$) are:
\begin{eqnarray}
\cI_r(\varPhi) &=& -\cI_r(\log C) \\
\cD_r(\varPhi) &=& \cD_r(\log C) \\
V_r &=& 0 \\
V_\theta &=& 4\log\pars{\frac{1 + \exp\left\{\cI_\theta(\zeta)/2\right\}}{2}} \cdot 
			\cD_\theta(\zeta) \\
  \zeta &=& - \log \gamma - \cI_r(\varPhi)
\end{eqnarray}
The equations are discretized by using ``ghost'' points, outside the interior of the domain:
\begin{eqnarray}
  \frac{\varPhi[0,j] + \varPhi[1,j]}{2} &=& 
	-\frac{\log C[0,j] + \log C[1,j]}{2} \\
  \frac{\varPhi[1,j] - \varPhi[0,j]}{\Delta r} &=& 
	\frac{\log C[1,j] - \log C[0,j]}{\Delta r} 
\end{eqnarray}
By adding and subtracting the equations, we have:
\begin{eqnarray}
\varPhi[0,j] &=& -\log\brcs{C[1,j]} \\
C[0,j] &=& \exp\brcs{-\varPhi[1,j]} \\
  \zeta[j] &=& \cV -\frac{\varPhi[0,j] + \varPhi[1,j]}{2} 
                     = -\log\gamma -\frac{\varPhi[0,j] + \varPhi[1,j]}{2}
\end{eqnarray}
The slipping conditions can be written by:
\begin{eqnarray}
V_r[0,j] &=& 0 \\
\frac{V_\theta[0,j] + V_\theta[1,j]}{2} &=& V_\theta[j] = 
4\log\pars{\frac{1 + \exp\left\{\frac{1}{2}\frac{\zeta[j] + \zeta[j+1]}{2}\right\}}{2}} 
\cdot \frac{\zeta[j+1] - \zeta[j]}{\bar{\theta}_{j+1} - \bar{\theta}_{j}} \\
  V_\theta[0,j] &=& 2 V_\theta[j] - V_\theta[1,j]
\end{eqnarray}

@@

Using an algebra of operators, the composition of operators is well defined:
\begin{eqnarray}
(\cO_1 \circ \cO_2)(\bx) &=& \cO_1(\cO_2(\bx))
\end{eqnarray}
Note that the differentiation of composed operators is computed using chain rule:
\begin{eqnarray}
\bnabla(\cO_1 \circ \cO_2)(\bx) &=& \bnabla \cO_1(\cO_2(\bx)) \cdot \bnabla \cO_2(\bx)
\end{eqnarray}

Thus, in order to solve the equation $\cO(\bx) = \bzero$, a Newton step can 
be computed by:
\begin{eqnarray}
\bzero = \cO(\bx_n + \Delta \bx_n) &\approx& \cO(\bx_n) + \bnabla \cO(\bx_n)\cdot\Delta\bx_n \\
\bnabla \cO(\bx_n) \cdot \Delta \bx_n &=& -\cO(\bx_n) \\
\bx_{n+1} &=& \bx_{n} + \Delta \bx_n 
\end{eqnarray}

Therefore, in order to apply Newton method, the residual vector $\cO(\bx)$ and
the gradient matrix of the operator $\bnabla \cO(\bx)$ are to be computed.
Then, a sparse linear system is to be solved to yield current step vector $\Delta \bx$.

When the method converges $\bx_n \rightarrow \bx_\infty$, 
the convergence is quadratic:
\begin{eqnarray}
\|\bx_{n+1} - \bx_\infty\| \le \|\bx_{n} - \bx_\infty\|^2
\end{eqnarray}
Thus, when the initial guess is close enough to the solution, the solver
will converge in very few steps.

\subsubsection{Operator Representation}
System variable $\bx$ is defined as the concatenation $[\,\varPhi, C, V_r, V_\theta, P\,]$,
taking values in the interior of the problem domain grid.

Thus, each problem variable can be computed by applying an appropriate projection operator:
\begin{eqnarray}
X &=& \cP_X(\bx) \mbox{ for } X \in \left\{\varPhi, C, V_r, V_\theta, P\right\}
\end{eqnarray}

The extended variables $\tilde X$ (with ``ghost'' points values 
from boundary conditions) are be computed, by applying the following non-linear operators:
\begin{eqnarray}
\tilde{\varPhi} &=& \cB_\varPhi(\varPhi, C; \beta) \\
\tilde{C} &=& \cB_C(C, \varPhi) \\
\tilde{V_r} &=& \cB_{V_r}(V_r; \cU) \\
\tilde{V_\theta} &=& \cB_{V_\theta}(V_\theta, C, \varPhi; \cU, \gamma) \\
\tilde{P} &=& P
\end{eqnarray}

The concatenated equations of the system (as described above) can be written 
as a non-linear operator $\cE(\tilde\bx) = \bzero$ acting on the extended variables $\tilde\bx = \cB(\bx)$:
The problem can be casted as solving $\cO(\bx) = \bzero$ for $\cO = \cE \circ \cB$,
where $\cB$ is the ``ghost'' point extension operator (derived using boundary conditions) 
and $\cE$ is the conservation equations operator (derived by concatenation of the system
equations).

\subsubsection{Steady-state}
After convergence, the ``ghost'' points values are updated using $\cB$,
so the force $\tT \cdot \bn$ acting on the ion-exchanger surface
can be computed.

Due to symmetry considerations, the total force $\bF$ is 
aligned with $\ui$ and it vanishes iff $\ui \cdot \bF = 0$.
\begin{eqnarray}
F_\imath = \ui \cdot \bF &=& \oint_\mathcal{S} \ui \cdot \tT \cdot \bnhat  dA = 
\int_0^\pi f_\imath(\theta) \cdot 2\pi \sin\theta d\theta 
\\  
f_\imath &=& \imath \cdot \tT \cdot \bnhat
\\  
f_\imath &=& \pars{-P + 2\cD_r(V_r) + 
\frac{1}{2}\pars{\cD_r(\varPhi)}^2 - \frac{1}{2r^2}\pars{\cD_\theta(\varPhi)}^2}\cos\theta 
\\  
&& -\pars{\cD_r(V_\theta) - \frac{V_\theta}{r}
+ \frac{1}{r}\cD_r(\varPhi) \cD_{\theta}(\varPhi)}\sin\theta \\
 \mathcal{S} &=& \{\br : \|\br\|_2 = 1\}
\end{eqnarray}

The integral above is approximated by 1D numerical quadrature, to yield 
the total force $F_\imath$ as a function of the drift velocity $\cU$.

@@

\subsubsection{Operator Interface}
The base \verb|Operator| interface is defined by 
supporting the computation of a residual/result vector \verb|op.res()|
and a gradient matrix \verb|op.grad()|.
This interface is implemented in various operators classes,
which are used to constuct $\cO = \cE \circ \cB$ and to apply Newton method
to the problem by computing the residual and the gradient automatically.
Each such operator may be defined using other operators, so its residual
and gradient are computed recursively.

\subsubsection{Unary operators}
\paragraph{Linear operator}
If $\cO$ is linear, it can be represented by a matrix $L$, such that:
\begin{eqnarray}
\cO(\bx) &=& L \bx \\
\bnabla\cO &=& L 
\end{eqnarray}
Note that finite difference $\cD$, interpolation $\cI$ and 
upwind selection $\mathcal{U}$ operators
can be implemented as linear sparse operators, by constructing 
an appropriate sparse matrix $L$, having $O(\dim \bx)$ non-zeroes)

@@

\subsubsection{Binary operators}
Let $\cO_1$ and $\cO_2$ be two operators, so their pointwise addition, subtraction
and multiplication are defined as follows:
\begin{eqnarray}
(\cO_1 \pm \cO_2)(\bx) &=& \cO_1)(\bx) \pm \cO_2(\bx) \\
\bnabla (\cO_1 \pm \cO_2)(\bx)   &=& \bnabla\cO_1(\bx) \pm \bnabla\cO_2(\bx) \\
(\cO_1 \cdot \cO_2)(\bx) &=& \diag(\diag(\cO_1(\bx)) \cdot \diag(\cO_2(\bx))) \\
\bnabla (\cO_1 \cdot \cO_2)(\bx) &=& \diag(\cO_2(\bx)) \bnabla\cO_1(\bx) 
                                  + \diag(\cO_1(\bx)) \bnabla\cO_2(\bx)
\end{eqnarray}

\subsubsection{N-ary operators}
Given $N$ operators $\{\cO_i\}_{i=1}^N$, their concatenation $\cO$ is defined as:
\begin{eqnarray}
\cO(\bx) &=& [\cO_1(\bx); \ldots; \cO_N(\bx)] \\
(\bnabla\cO)(\bx) &=& [(\bnabla\cO_1)(\bx); \ldots; (\bnabla\cO_N)(\bx)]
\end{eqnarray}

\subsubsection{Iterative Newton solver}
Given an initial solution $\bx_0$, we apply the following algorithm 
until convergence ($\|\Delta \bx\| < \eps$):
\begin{enumerate}
\item Update the system: $\verb|op| = \verb|System|(\bx_{n})$
\item Compute the gradient: $\boldsymbol{A}_n = \verb|op.grad()|$
\item Compute the right-hand side: $\boldsymbol{b}_n = \verb|-op.res()|$
\item Solve sparse linear system: $\boldsymbol{A}_n \Delta \bx_n = \boldsymbol{b}_n$
\item Update solution: $\bx_{n+1} = \bx_{n} + \Delta \bx_{n}$
\end{enumerate}

\subsubsection{Total force integration -- $F_\imath$}
After the solver converges ($\bx_n \rightarrow \bx_\infty$), 
the ``ghost'' points values are computed by applying $\cB$ and
solving for zero force  $\brhat$ component on $r=1$ boundary

@@

\subsubsection{Steady state drift velocity -- $\cU$}
In steady state, the force acting on the particle $F_\imath$ shoud vanish.
This yields an equation on $\cU$:
\begin{eqnarray}
F_\imath(\cU) &=& 0
\end{eqnarray}
This equation can be solved using 1D root finding algorithm, e.g. Secant method:
\begin{eqnarray}
\cU_{n+1} = \frac{F_\imath(\cU_{n})\cU_{n-1} - F_\imath(\cU_{n-1})\cU_{n}}
{F_\imath(\cU_{n}) - F_\imath(\cU_{n-1})} \rightarrow \cU_\infty
\end{eqnarray}
The result $\cU_\infty$ is the 
steady state particle drift velocity for the electric field $\beta$.

>>>>

The boundary conditions at the ion-exchange boundary ($r=1$) are:
\begin{eqnarray}
\cI_r(\varPhi) &=& -\cI_r(\log C) \\
\cD_r(\varPhi) &=& \cD_r(\log C) \\
V_r &=& 0 \\
V_\theta &=& 4\log\pars{\frac{1 + \exp\left\{\cI_\theta(\zeta)/2\right\}}{2}} \cdot 
			\cD_\theta(\zeta) \\
  \zeta &=& - \log \gamma - \cI_r(\varPhi)
\end{eqnarray}
The equations are discretized by using ``ghost'' points, outside the interior of the domain:
\begin{eqnarray}
 \varPhi[1/2, j] = \frac{\varPhi[0,j] + \varPhi[1,j]}{2} &=& 
	-\frac{\log C[0,j] + \log C[1,j]}{2} = -\log C[1/2, j] \\
  \frac{\varPhi[1,j] - \varPhi[0,j]}{\Delta r} &=& 
	\frac{\log C[1,j] - \log C[0,j]}{\Delta r} 
\end{eqnarray}
By adding and subtracting the equations, we have:
\begin{eqnarray}
\varPhi[0,j] &=& -\log\brcs{C[1,j]} \\
C[0,j] &=& \exp\brcs{-\varPhi[1,j]} \\
  \zeta[j] &=& \cV - \varPhi[1/2,j]
                     = -\log\gamma -\frac{\varPhi[0,j] + \varPhi[1,j]}{2}
\end{eqnarray}
The slipping conditions can be written by:
\begin{eqnarray}
V_r[0,j] &=& 0 \\
V_\theta[1/2,j] &=& 
4\log\pars{\frac{1 + \exp\left\{\frac{1}{2}\frac{\zeta[j] + \zeta[j+1]}{2}\right\}}{2}} 
\cdot \frac{\zeta[j+1] - \zeta[j]}{\bar{\theta}_{j+1} - \bar{\theta}_{j}} \\
  V_\theta[0,j] &=& 2 V_\theta[1/2, j] - V_\theta[1,j]
\end{eqnarray}

@@

Using an algebra of operators, the composition of operators is well defined:
\begin{eqnarray}
(\cO_1 \circ \cO_2)(\bx) &=& \cO_1(\cO_2(\bx))
\end{eqnarray}
Note that the differentiation of composed operators is computed by using the chain rule:
\begin{eqnarray}
\bnabla(\cO_1 \circ \cO_2)(\bx) &=& \bnabla \cO_1(\cO_2(\bx)) \cdot \bnabla \cO_2(\bx)
\end{eqnarray}

In order to solve the equation $\cO(\bx) = \bzero$, the Newton's method can 
be applied, given an initial solution $\bx_0$:
\begin{eqnarray}
\bzero = \cO(\bx_n + \Delta \bx_n) &\approx& \cO(\bx_n) + \bnabla \cO(\bx_n)\cdot\Delta\bx_n \\
\bnabla \cO(\bx_n) \cdot \Delta \bx_n &=& -\cO(\bx_n) \\
\bx_{n+1} &=& \bx_{n} + \Delta \bx_n 
\end{eqnarray}

Newton method step required the computation of the residual vector $\cO(\bx)$ and
the gradient matrix of the operator $\bnabla \cO(\bx)$, 
given the current solution $\bx = \bx_n$.
Then, a sparse linear system is to be solved to yield the update for the soltion vector
$\Delta \bx_n = \bx_{n+1} - \bx_n$.

When the method does converges $\bx_n \rightarrow \bx_\infty$, 
the convergence is quadratic:
\begin{eqnarray}
\|\bx_{n+1} - \bx_\infty\| \le k \|\bx_{n} - \bx_\infty\|^2
\end{eqnarray}
Thus, when the initial guess is close enough to the solution, the solver
will converge in very few steps.

\subsubsection{Operator Representation}
System variable $\bx$ is defined as the concatenation $[\,\varPhi, C, V_r, V_\theta, P\,]$,
taking values in the interior of the problem domain grid.
So each problem variable can be computed by applying an appropriate projection operator
on $\bx$:
\begin{eqnarray}
X &=& \cP_X(\bx) \mbox{ for } X \in \left\{\varPhi, C, V_r, V_\theta, P\right\}
\end{eqnarray}

The extended variables $\tilde X$ (with ``ghost'' points values 
from boundary conditions) are be computed, by applying the following non-linear operators:
\begin{eqnarray}
\tilde{\varPhi} &=& \cB_\varPhi(\varPhi, C; \beta) \\
\tilde{C} &=& \cB_C(C, \varPhi) \\
\tilde{V_r} &=& \cB_{V_r}(V_r; \cU) \\
\tilde{V_\theta} &=& \cB_{V_\theta}(V_\theta, C, \varPhi; \cU, \gamma) \\
\tilde{P} &=& P
\end{eqnarray}
The extended system variable can be written as $\tilde{\bx} = \cB(\bx)$, 
where $\cB$ is the ``ghost'' point extension operator (derived using boundary conditions)
so the concatenated equations of the system (as described above) can be written 
as a non-linear operator $\cE(\tilde\bx) = \bzero$ acting on the extended system variable,
where $\cE$ is the conservation equations operator (derived by concatenation of the system
equations).

The problem can be casted as to solve the following non-linear system:
\begin{eqnarray}
\bzero = \cE (\tilde{\bx}) = \cE (\cB (\bx)) = \cO(\bx)
\end{eqnarray}
 where $\cO = \cE \circ \cB$.

\subsubsection{Steady-state}
Assume that $\cU \ui$ is the velocity of the ion exchanger 
(given the fluid is at rest far away from the particle).
After the convergence of the Newton method solver, $\cO(\bx_n) \rightarrow \bzero$, 
the ``ghost'' points values are updated using $\tilde\bx = \cB(\bx_\infty)$,
so the surface force $\boldsymbol{f} = \tT \cdot \bn$ acting on the ion-exchanger surface
can be computed by integration on the ion exchanger's surface $\mathcal S$.

Due to symmetry considerations, the total force $\bF$ is 
aligned with $\ui$ and it vanishes iff $\ui \cdot \bF = 0$.
\begin{eqnarray}
\bF &=& \oint_\mathcal{S} \boldsymbol{f}  dA = 
\oint_\mathcal{S} \tT \cdot \bnhat  dA \\
F_\imath = \ui \cdot \bF &=& \oint_\mathcal{S} \ui \cdot \boldsymbol{f}  dA = 
\oint_\mathcal{S} \ui \cdot \tT \cdot \bnhat  dA = 
\int_0^\pi f_\imath(\theta) \cdot 2\pi \sin\theta d\theta 
\\  
f_\imath &=& \ui \cdot \tT \cdot \brhat
\\  
f_\imath &=& \pars{-P + 2\cD_r(V_r) + 
\frac{1}{2}\pars{\cD_r(\varPhi)}^2 - \frac{1}{2r^2}\pars{\cD_\theta(\varPhi)}^2}\cos\theta 
\\  
&& -\pars{\cD_r(V_\theta) - \frac{V_\theta}{r}
+ \frac{1}{r}\cD_r(\varPhi) \cD_{\theta}(\varPhi)}\sin\theta \\
 \mathcal{S} &=& \{\br : \|\br\|_2 = 1\}
\end{eqnarray}

The integral for $F_\imath$ is approximated by 1D numerical quadrature, to yield 
the total force $F_\imath(\cU)$ as a function of the drift velocity $\cU$.

@@

\subsubsection{Operator Interface $ \rightarrow \cO$}
The base \verb|Operator| interface is defined by 
supporting the computation of a residual vector \verb|op.res()|
and a gradient matrix \verb|op.grad()|, given an input vector $\bx$.

This interface is implemented by specific operators classes,
which are used to constuct $\cO = \cE \circ \cB$ operator, so that 
Newton method solver (that requires computing the residual and the gradient) 
can be applied automatically.

Each such operator may be have other operators as its inputs, so the residual
and the gradient are computed recursively.

\paragraph{Linear operator}
If $\cO$ is linear, it can be represented by a matrix $L$, such that:
\begin{eqnarray}
\cO(\bx) &=& L \bx \\
\bnabla\cO &=& L 
\end{eqnarray}
Note that finite difference $\cD$, interpolation $\cI$ and 
upwind selection $\mathcal{U}$ operators
can be implemented as linear sparse operators, by constructing 
an appropriate sparse matrix $L$, having $O(\dim \bx)$ non-zeroes)

@@

\paragraph{Binary operators}
Let $\cO_1$ and $\cO_2$ be two operators, so their pointwise addition, subtraction
and multiplication are defined as follows:
\begin{eqnarray}
(\cO_1 \pm \cO_2)(\bx) &=& \cO_1)(\bx) \pm \cO_2(\bx) \\
\bnabla (\cO_1 \pm \cO_2)(\bx)   &=& \bnabla\cO_1(\bx) \pm \bnabla\cO_2(\bx) \\
(\cO_1 \cdot \cO_2)(\bx) &=& \diag(\diag(\cO_1(\bx)) \cdot \diag(\cO_2(\bx))) \\
\bnabla (\cO_1 \cdot \cO_2)(\bx) &=& \diag(\cO_2(\bx)) \bnabla\cO_1(\bx) 
                                  + \diag(\cO_1(\bx)) \bnabla\cO_2(\bx)
\end{eqnarray}

\paragraph{N-ary operators}
Given $N$ operators $\{\cO_i\}_{i=1}^N$, their concatenation $\cO$ is defined as:
\begin{eqnarray}
\cO(\bx) &=& [\cO_1(\bx); \ldots; \cO_N(\bx)] \\
(\bnabla\cO)(\bx) &=& [(\bnabla\cO_1)(\bx); \ldots; (\bnabla\cO_N)(\bx)]
\end{eqnarray}

\subsubsection{Iterative Newton solver $\rightarrow \bx$}
Given an initial solution $\bx_0$ and specific values for $\beta$ and $\cU$, 
we apply the following algorithm 
until convergence ($\|\Delta \bx\| < \eps$):
\begin{enumerate}
\item Update the system: $\verb|op| = \verb|System|(\bx_{n})$
\item Compute the gradient: $\boldsymbol{A}_n = \verb|op.grad()|$
\item Compute the right-hand side: $\boldsymbol{b}_n = \verb|-op.res()|$
\item Solve sparse linear system: $\boldsymbol{A}_n \Delta \bx_n = \boldsymbol{b}_n$
\item Update solution: $\bx_{n+1} = \bx_{n} + \Delta \bx_{n}$
\end{enumerate}

\subsubsection{Total force integration $\rightarrow F_\imath$}
After the solver converges ($\bx_n \rightarrow \bx_\infty$), 
the ``ghost'' points values are computed by applying $\cB$ and
solving for zero force  $\brhat$ component on $r=1$ boundary

@@

\subsubsection{Steady state drift velocity $\rightarrow \cU$}
In steady state, the force acting on the particle $F_\imath$ shoud vanish.
This yields an equation on $\cU$:
\begin{eqnarray}
F_\imath(\cU) &=& 0
\end{eqnarray}
This equation can be solved using 1D root finding algorithm, e.g. Secant method:
\begin{eqnarray}
\cU_{n+1} = \frac{F_\imath(\cU_{n})\cU_{n-1} - F_\imath(\cU_{n-1})\cU_{n}}
{F_\imath(\cU_{n}) - F_\imath(\cU_{n-1})} \rightarrow \cU_\infty
\end{eqnarray}
The result $\cU_\infty$ is the 
steady state particle drift velocity for the electric field $\beta$.
